---
title: "Your Personal DeeDee Summary"
author: "DeeDee Summary Function"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: yes
    toc_float: yes
    theme: cosmo
    code_folding: show
    code_download: yes
  pdf_document:
    toc: yes
always_allow_html: yes
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  warning = FALSE,
  message = FALSE,
  error = TRUE
)
```

## About this document

This document contains the results of running all main DeeDee functions on your provided input data and parameters. 

Your contrasts:
```{r echo = TRUE}
dea(dde)
```


```{r echo = FALSE, results = FALSE}
library("DeeDee")
```


## DeeDee scatter plot

```{r echo = TRUE}
deedee_scatter(dde = dde,
               pthresh = pthresh,
               select1 = scatter_select1,
               select2 = scatter_select2,
               color_by = scatter_color_by)
```

Your parameters: 

* pthresh = `r pthresh` 
* select1 = `r scatter_select1`
* select2 = `r scatter_select2`
* color_by = `r scatter_color_by`

***

## DeeDee heatmap

```{r echo = TRUE}
deedee_heatmap(dde = dde,
               pthresh = pthresh,
               show_first = heatmap_show_first,
               show_gene_names = heatmap_show_gene_names,
               dist = heatmap_dist,
               clust = heatmap_clust,
               show_na = heatmap_show_na)
```

Your parameters: 

* pthresh = `r pthresh` 
* show_first = `r heatmap_show_first`
* show_gene_names = `r heatmap_show_gene_names`
* dist = `r heatmap_dist`
* clust = `r heatmap_clust`
* show_na = `r heatmap_show_na`

***

## DeeDee Venn diagram

```{r echo = TRUE}
deedee_venn(dde = dde,
            pthresh = pthresh,
            mode = venn_mode)
```

Your parameters: 

* pthresh = `r pthresh` 
* mode = `r venn_mode`

***

## DeeDee UpSet plot

```{r echo = TRUE}
deedee_upset(dde = dde,
             pthresh = pthresh,
             mode = upset_mode,
             min_setsize = upset_min_setsize)
```

Your parameters: 

* pthresh = `r pthresh` 
* mode = `r upset_mode`
* min_setsize = `r upset_min_setsize`

***

## DeeDee Q-Q plot

```{r echo = TRUE}
deedee_qqmult(dde = dde,
              pthresh = pthresh,
              ref = qqmult_ref)
```

Your parameters: 

* pthresh = `r pthresh`
* ref = `r qqmult_ref` 

***

## DeeDee CAT plot

```{r echo = TRUE}
deedee_cat(dde = dde,
           pthresh = pthresh,
           ref = cat_ref,
           maxrank = cat_maxrank,
           mode = cat_mode)
```

Your parameters: 

* pthresh = `r pthresh` 
* ref = `r cat_ref`
* maxrank = `r cat_maxrank`
* mode = `r cat_mode`

***

## Session info {-}

```{r}
sessionInfo()
```
