---
title: >
  The `DeeDee` User's Guide
author:
- name: Lea Rothörl
  affiliation: 
  - Institute of Medical Biostatistics, Epidemiology and Informatics (IMBEI), Mainz
- name: Federico Marini
  affiliation: 
  - Institute of Medical Biostatistics, Epidemiology and Informatics (IMBEI), Mainz
  - Research Center for Immunotherapy (FZI), Mainz
  email: marinif@uni-mainz.de
date: "`r BiocStyle::doc_date()`"
package: "`r BiocStyle::pkg_ver('DeeDee')`"
output: 
  BiocStyle::html_document:
    toc_float: true
    code_folding: show
    code_download: yes
vignette: >
  %\VignetteIndexEntry{The DeeDee User's Guide}
  %\VignetteEncoding{UTF-8}  
  %\VignettePackage{DeeDee}
  %\VignetteKeywords{GeneExpression, RNASeq, Sequencing, Visualization, QualityControl, GUI}
  %\VignetteEngine{knitr::rmarkdown}
editor_options: 
  chunk_output_type: console
bibliography: DeeDee_bibliography.bib
---

<style type="text/css">
.smaller {
  font-size: 10px
}
</style>

**Compiled date**: `r Sys.Date()`

**Last edited**: 2023-01-12

**License**: `r packageDescription("DeeDee")[["License"]]`

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment  = "#>",
  error    = FALSE,
  warning  = FALSE,
  eval     = TRUE,
  message  = FALSE,
  fig.width = 10
)

options(width = 100)
```

<hr>

# Introduction {#introduction}

This vignette describes how to use the `r BiocStyle::Biocpkg("DeeDee")` package for ......

This package provides ...... 

In order to use `r BiocStyle::Biocpkg("DeeDee")` in your workflow, ...... the following inputs are required: 

In the remainder of this vignette, we will illustrate the main features of `r BiocStyle::Biocpkg("DeeDee")` on a publicly available dataset from Alasoo, et al. "Shared genetic effects on chromatin and gene expression indicate a role for enhancer priming in immune response", published in Nature Genetics, January 2018 [@Alasoo2018]
[doi:10.1038/s41588-018-0046-7](https://doi.org/10.1038/s41588-018-0046-7).

The data is made available via the `r BiocStyle::Biocpkg("macrophage")` Bioconductor package, which contains the files output from the Salmon quantification (version 0.12.0, with Gencode v29 reference), as well as the values summarized at the gene level, which we will use to exemplify.

In the `macrophage` experimental setting, the samples are available from 6 different donors, in 4 different conditions (naive, treated with Interferon gamma, with SL1344, or with a combination of Interferon gamma and SL1344).

......

We will focus our attention on the following comparisons:

* between Interferon gamma treated samples vs naive samples.
......

# Getting started {#gettingstarted}

To install this package, start R and enter:

```{r install, eval=FALSE}
if (!requireNamespace("BiocManager", quietly = TRUE))
  install.packages("BiocManager")

BiocManager::install("DeeDee")
```

Once installed, the package can be loaded and attached to your current workspace as follows:

```{r loadlib, eval = TRUE}
library("DeeDee")
```

If you have all four input elements ready, 

......

you can launch the `GeneTonic()` app by running:

......


# The `DeeDeeExperiment` class

yadda yadda

......

# A typical analysis workflow with `DeeDee` {#typicalworkflow}


a number of functions which can be used also as standalone components for a workflow.



# Using `DeeDee` interactively: the `deedee_app()`


Description of the `DeeDee` user interface {#userinterface}

What it does

How to use it




# Additional Information {#additionalinfo}

Bug reports can be posted on the Bioconductor support site (https://support.bioconductor.org/) or raised as issues in the `DeeDee` GitHub repository
......
(https://github.com/federicomarini/GeneTonic/issues). 

The GitHub repository also contains the development version of the package, where new functionality is added over time - careful, you might be running bleeding edge versions!

The authors appreciate well-considered suggestions for improvements or new features, or even better, pull requests.

If you use `DeeDee` for your analysis, please cite it as shown below:

```{r cite}
citation("DeeDee")
```


# FAQs {#faqs}

**Q: Why DeeDee?**

A: For this and that

**Q: I'd like to try `DeeDee` but I could not install it/I just want a quick peek into it. What can I do for this?**

A: We set up an instance of `DeeDee` running on the `macrophage` dataset at this address: http://shiny.imbei.uni-mainz.de:3838/DeeDee.

# Session Info {- .smaller}

```{r sessioninfo}
sessionInfo()
```

# References {-}

```{r}
```















```{r, echo = FALSE, warning = FALSE, message = FALSE, results = "hide"}
library(DeeDeeLegacy)
data(DE_results_IFNg_naive, package = "DeeDee")
IFNg_naive <- deedee_prepare(IFNg_naive, "DESeq2")
data(DE_results_IFNg_both, package = "DeeDee")
IFNg_both <- deedee_prepare(IFNg_both, "DESeq2")
data(DE_results_Salm_naive, package = "DeeDee")
Salm_naive <- deedee_prepare(Salm_naive, "DESeq2")
data(DE_results_Salm_both, package = "DeeDee")
Salm_both <- deedee_prepare(Salm_both, "DESeq2")

DeeDee_obj <- list(IFNg_naive = IFNg_naive, 
                   IFNg_both = IFNg_both, 
                   Salm_naive = Salm_naive, 
                   Salm_both = Salm_both)
```

# What DeeDee is for

When you want to compare results from multiple Differential Expression Analyses (DEAs), the DeeDee package is your friend. It contains various functions, and a Shiny App combining them all, that help shed light on the similarities and differences between the experiments in question. DeeDee is designed to be used after the application of a DE analysis program (like DESeq2, edgeR or limma) on the single DEAs. 


# How to Install

You can install the current development version from GitHub:
```
library("remotes")
remotes::install_github("lea-rothoerl/DeeDee", 
                        dependencies = TRUE, build_vignettes = TRUE)
```


# Example Dataset Used in this Vignette

The examples in the following chapters utilize the data from the Bioconductor package `macrophage`([Human macrophage immune response](https://bioconductor.org/packages/release/data/experiment/html/macrophage.html "Bioconductor - macrophage")). The macrophage dataset includes data from 24 RNA-seq samples of human macrophages exposed to different conditions: naive, associated with IFNg (an interferon), SL1344 (a strain of Salmonella), or both. The preprocessing of the data was done using DESeq2, obtaining the following four DEAs: naive vs. IFNg, IFNg vs. both, naive vs. Salmonella, and Salmonella vs. both. 


# What to Provide

The main DeeDee functions work on tables of logFC- ("logFC") and p-values ("pval") from DEAs, with the gene identifiers as row names. This table can be built by the user manually or by using `deedee_prepare`. The `deedee_prepare` function accepts results from three of the most used R DE analysis packages (DESeq2, edgeR, limma) and converts them to digestible tables. Which of these packages was used to analyze the raw data needs to be specified using the parameter `input_type`. If, for example, you have a DESeq2 result called `DESeq2_res` and want to convert it to a DeeDee table, you can use the following code chunk to do so. 

```{r, warning = FALSE, message = FALSE, eval = FALSE}
inp <- deedee_prepare(data = DESeq2_res, input_type = "DESeq2")
```

Every [DeeDee main function](#main-functions) needs to be fed with a (named) list of at least two of these tables as input (`data`). A threshold for p-values can be specified with the parameter `pthresh` (default = 0.05). Besides these, most functions have additional parameters that will be explained in the respective sections below. All functions produce colorblind-friendly output.

For our example, the `deedee_prepare` results are `inp1` - `inp4` (naive vs. IFNg, IFNg vs. both, naive vs. Salmonella, and Salmonella vs. both from [macrophage](#example-dataset-used-in-this-vignette)). The named input list is created as follows. 

```{r, warning = FALSE, message = FALSE, results = "hide", eval = FALSE}
DeeDee_obj <- list(naive_IFNg = inp1, IFNg_both = inp2, naive_Salm = inp3, Salm_both = inp4)
```


## DESeq2

When `DESeq2` is applied to analyze the raw data, `deedee_prepare` accepts the output from the `results()` function. 


## edgeR

When `edgeR` is applied to analyze the raw data, `deedee_prepare` accepts the output from the `topTable()` function. 


## limma

When `limma` is applied to analyze the raw data, `deedee_prepare` accepts the output from the `exactTest()` function. 


\pagebreak
# Exported Functions
## DeeDee Scatter Plot

The function `deedee_scatter()` creates a scatterplot of logFC values of the genes in two input datasets. If the input DeeDee list contains more than two datasets (like our example `DeeDee_obj` does),  the `select1` (default = 1) and `select2` (default = 2) parameters specify which ones will be plotted (selected by list index). `deedee_scatter()` includes a `color_by` parameter that can be set to "pval1" (default) or "pval2". The points will be colored according to the color scheme given on the right side of the output. 

```{r, echo = TRUE, warning = FALSE, message = FALSE}
deedee_scatter(data = DeeDee_obj, select1 = 2, select2 = 3, color_by = "pval1", pthresh = 0.05)
```


\pagebreak
## DeeDee Heatmap

The function `deedee_heatmap()` creates a heatmap of the logFC values for all common genes in every input dataset. The color key is given on the right side of the plot. Additionally to the standard parameters explained in [Input](#what-to-provide), the function takes a numeric `show_first` value (default = 25), specifying the number of genes depicted (if the total number of genes is smaller than `show_first`, all genes are shown). It also digests a logical `show_gene_names` value (default = FALSE) that determines if the gene identifiers (rownames in `deedee_prepare` results) will be displayed in the heatmap, and a logical `show_na`, defining if genes with NAs (in less than half of the contrasts) are included. The distance measure (`dist`, values: `euclidean` (default), `manhattan`, `pearson`, `spearman`) and clustering method (`clust`, values: `single`, `complete`, `average` (default), `centroid`) can be chosen as well. 

```{r, echo = TRUE, warning = FALSE, message = FALSE}
deedee_heatmap(data = DeeDee_obj, show_first = 25, show_gene_names = FALSE, dist = "manhattan", clust = "centroid", show_na = FALSE, pthresh = 0.05)
```

Because the heatmap is produced with the [InteractiveComplexHeatmap](https://bioconductor.org/packages/release/bioc/html/InteractiveComplexHeatmap.html) package, a Shiny window with the heatmap and entailed interactivity can be opened by executing the command `InteractiveComplexHeatmap::ht_shiny(res)` to a result `res` of `deedee_heatmap()`. For more information, please refer to the [InteractiveComplexHeatmap Vignette](https://bioconductor.org/packages/release/bioc/vignettes/InteractiveComplexHeatmap/inst/doc/InteractiveComplexHeatmap.html).


\pagebreak
## DeeDee Venn Diagram

The function `deedee_venn()` creates a Venn diagram depicting the overlaps of differentially expressed genes in the input datasets. To keep the Venn diagram easy on the eye, the `data` list may contain no more than four datasets. To compare more DEAs in a similar, set-based manner, please make use of [`deedee_upSet()`](#deedee-upset-plot). The parameter `mode` can be set to `up`, `down`, or `both` (default), specifying if only up-regulated, down-regulated or both DE genes will be counted. 

```{r, echo = TRUE, warning = FALSE, message = FALSE}
# deedee_venn(data = DeeDee_obj, mode = "both", pthresh = 0.05)
```


\pagebreak
## DeeDee UpSet Plot

The function `deedee_upset()` creates an UpSet plot depicting the overlaps of differentially expressed genes in the input datasets. Contrary to [`deedee_venn()`](#deedee-venn-diagram), the UpSet plot can compare multiple (and even more than four) datasets in a visually pleasing way. The parameter `mode` can be set to `up`, `down`, `both`, or `both_colored` (default), specifying if only up-regulated, down-regulated or both DE genes shall be depicted. If `both_colored` is chosen, the result is the same UpSet plot as for `both`, but the shares of the intersections where all samples have positive/negative logFC values are colored according to the color key. The minimum size for an intersection to be included in the plot can be defined via the parameter `min_setsize`, default is 10.

```{r, echo = TRUE, warning = FALSE, message = FALSE}
deedee_upset(data = DeeDee_obj, mode = "both_colored", min_setsize = 15, pthresh = 0.05)
```



\pagebreak
## DeeDee Q-Q Plot

The function `deedee_qq()` compares the statistical distributions of two input datasets. If the input data list contains more than two datasets (like our example `DeeDee_obj` does),  the `select1` (default = 1) and `select2` (default = 2) parameters specify which ones are to be used. If the resulting curve resembles a straight line with a slope of 1, the distributions are similar (perfect straight line = identical distributions). `deedee_qq()` includes a `color_by` parameter that can be set to `pval1` (default) or `pval2`. The points generating the curve will be colored according to the color key displayed on the right of the output. 

```{r, echo = TRUE, warning = FALSE, message = FALSE}
deedee_qq(data = DeeDee_obj, select1 = 1, select2 = 3, color_by = "pval2", pthresh = 0.05)
```

The function `deedee_qqmult()` makes the same calculations as `deedee_qq`, but it is capable of depicting multiple Q-Q lines of different contrasts against the same reference in one plot. The curves are then colored by contrast, a coloration by p-value, like what `deedee_qq()` allows for, is not possible with `deedee_qqmult()`.

```{r, echo = TRUE, warning = FALSE, message = FALSE}
deedee_qqmult(data = DeeDee_obj, ref = 1, pthresh = 0.05)
```


\pagebreak
## DeeDee CAT Plot

The function `deedee_cat()` creates a plot depicting Concordance At the Top curves for a reference contrast (parameter `ref`, select contrast by list index). For each contrast except the reference, a curve indicating the concordance of the top genes in the contrast's logFC-sorted gene list, against the reference's, is displayed. The argument `mode` specifies the way of sorting the genes: by highest (`up`, default), lowest (`down`) or highest absolute (`both`) logFC value. The highest rank for which the concordance is calculated is given by the `maxrank` (default = 1000) parameter. 

```{r, echo = TRUE, warning = FALSE, message = FALSE}
deedee_cat(data = DeeDee_obj, ref = 2, maxrank = 800, pthresh = 0.05)
```

\pagebreak
## DeeDee Summary

`deedee_summary()` is the function to combine the results from the other DeeDee functions as a HTML document in a report-like manner. The function entails arguments to set the parameters for all included functions (for more information see the manual). The summary will be saved to a path specified with the argument `output_path` (default = "DeeDee_Summary.html" in the working directory). The parameter `overwrite` decides if a potentially existing document at the designated location will be overwritten (default = TRUE). Param `silent` can suppress messages (default = FALSE) and `open_file` decides if the resulting document will be opened (default = TRUE).

```{r, echo = TRUE, warning = FALSE, message = FALSE, eval = FALSE}
deedee_summary(deedee_list = DeeDee_obj,
               output_path = "DeeDee_Summary.html",
               overwrite = FALSE,
               pthresh = 0.05,
               scatter_select1 = 1,
               scatter_select2 = 2,
               scatter_color_by = "pval1",
               heatmap_show_first = 25,
               heatmap_show_gene_names = FALSE,
               heatmap_dist = "euclidean",
               heatmap_clust = "average",
               heatmap_show_na = FALSE,
               venn_mode = "both",                           
               upset_mode = "both_colored",
               upset_min_setsize = 10,
               qqmult_ref = 1,
               cat_ref = 1,
               cat_maxrank = 1000,
               cat_mode = "up",
               silent = FALSE,
               open_file = TRUE))
```


\pagebreak
## DeeDee Shiny App

Besides the standalone functions, the DeeDee package also contains an interactive Shiny web application. To open it, run the following command:

```{r, echo = TRUE, warning = FALSE, message = FALSE, eval = FALSE}
deedee_app()
```
It can work on a diverse range of input files (.txt, .xlsx, .RDS) with different contents (single DeeDee tables, lists of DeeDee tables, raw DEA result objects), as described in the `INFO` panel in the input tab. 
The App entails one tab for each of the above presented functions. The respective parameters can be set via self-explanatory input boxes in each tab. Additional interactive functionality, namely the [interactive heatmap](#deedee-heatmap), a brushing (area selection) of points in the Q-Q and scatter plot and the possibility to run an over-representation on the brushed points in the scatter plot are implemented.
More information on the usage can be found in the `INFO` panel in each respective tab.

Optionally, you can include a list of DeeDee tables as an argument to the `deedee_app(data = DeeDee_obj)` function. The input can be combined with uploads and used just like these. 

 
